\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-29 paper.qmd
tweaking the Background paragraph

<!-- - Connection between NLDR and MDS-->
Historically, $k-D$ representations of $p-D$ data have been computed using multidimensional Scaling (MDS) [@borg2005], which includes principal components analysis (PCA) [@jolliffe2011] as a special case.  The $k-D$ representation can be considered to be a layout of points in $k-D$ produced by an embedding procedure that maps the data from $p-D$. In MDS, the $k-D$ layout is constructed by minimizing a stress function that differences distances between points in $p-D$ with potential distances between points in $k-D$. Various formulations of the stress function result in non-metric scaling [@kruskal1964] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

@@

At first glance, thinking of of NLDR as a model fitted to the data might seem strange. It is a model in the sense that it is a "a simplified representation or abstraction of a system, process, or phenomenon in the real world" where the $p-D$ observations are the realization. From a statistical perspective we can consider the distances between points in the $k-D$ layout to be variance that the model explains, and the (relative) difference with their distances in $p-D$ is the error, or unexplained variance. Abstractly, we can also imagine that the positioning of points in 2D represents fitted values, that will have some prescribed position in $p-D$ that can be compared with the observed value.

>>>>

<!-- - Connection between NLDR and MDS-->
Historically, $k-D$ representations of $p-D$ data have been computed using multidimensional scaling (MDS) [@borg2005], which includes principal components analysis (PCA) [@jolliffe2011] as a special case.  The $k-D$ representation can be considered to be a layout of points in $k-D$ produced by an embedding procedure that maps the data from $p-D$. In MDS, the $k-D$ layout is constructed by minimizing a stress function that differences distances between points in $p-D$ with potential distances between points in $k-D$. Various formulations of the stress function result in non-metric scaling [@kruskal1964] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

@@

At first glance, thinking of of NLDR as a model fitted to the data might seem strange. It is a model in the sense that it is a "a simplified representation or abstraction of a system, process, or phenomenon in the real world". The $p-D$ observations are the realization of the phenomenon, and the $k-D$ NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the $k-D$ layout to be variance that the model explains, and the (relative) difference with their distances in $p-D$ is the error, or unexplained variance. Abstractly, we can also imagine that the positioning of points in 2D represents fitted values, that will have some prescribed position in $p-D$ that can be compared with the observed value.

### 2024-03-29 paper.tex
tweaking the Background paragraph

Historically, \(k-D\) representations of \(p-D\) data have been computed
using multidimensional Scaling (MDS) \citep{borg2005}, which includes
principal components analysis (PCA) \citep{jolliffe2011} as a special
case. The \(k-D\) representation can be considered to be a layout of
points in \(k-D\) produced by an embedding procedure that maps the data
from \(p-D\). In MDS, the \(k-D\) layout is constructed by minimizing a
stress function that differences distances between points in \(p-D\)
with potential distances between points in \(k-D\). Various formulations
of the stress function result in non-metric scaling \citep{kruskal1964}
and isomap \citep{silva2002}. Challenges in working with
high-dimensional data, including visualization, are outlined in
\citet{johnstone2009}.

@@

At first glance, thinking of of NLDR as a model fitted to the data might
seem strange. It is a model in the sense that it is a ``a simplified
representation or abstraction of a system, process, or phenomenon in the
real world'' where the \(p-D\) observations are the realization. From a
statistical perspective we can consider the distances between points in
the \(k-D\) layout to be variance that the model explains, and the
(relative) difference with their distances in \(p-D\) is the error, or
unexplained variance. Abstractly, we can also imagine that the
positioning of points in 2D represents fitted values, that will have
some prescribed position in \(p-D\) that can be compared with the
observed value.

>>>>

Historically, \(k-D\) representations of \(p-D\) data have been computed
using multidimensional scaling (MDS) \citep{borg2005}, which includes
principal components analysis (PCA) \citep{jolliffe2011} as a special
case. The \(k-D\) representation can be considered to be a layout of
points in \(k-D\) produced by an embedding procedure that maps the data
from \(p-D\). In MDS, the \(k-D\) layout is constructed by minimizing a
stress function that differences distances between points in \(p-D\)
with potential distances between points in \(k-D\). Various formulations
of the stress function result in non-metric scaling \citep{kruskal1964}
and isomap \citep{silva2002}. Challenges in working with
high-dimensional data, including visualization, are outlined in
\citet{johnstone2009}.

@@

At first glance, thinking of of NLDR as a model fitted to the data might
seem strange. It is a model in the sense that it is a ``a simplified
representation or abstraction of a system, process, or phenomenon in the
real world''. The \(p-D\) observations are the realization of the
phenomenon, and the \(k-D\) NLDR layout is the simplified
representation. From a statistical perspective we can consider the
distances between points in the \(k-D\) layout to be variance that the
model explains, and the (relative) difference with their distances in
\(p-D\) is the error, or unexplained variance. Abstractly, we can also
imagine that the positioning of points in 2D represents fitted values,
that will have some prescribed position in \(p-D\) that can be compared
with the observed value.
